A sample workflow

### 2025-04-29 analyses/african_mammals_workflow.qmd
Last edits for the day, I am done for now with workflow.

At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one sigma for all populations (CHECK HERE: do we really use one sigma for all populations?) and potentially our linearity assumption. One possibility would be to move to a non-linear regression, and use a more flexible model that would allow population numbers to stabilize, such as shown using a Gompertz model for the wildbeest dataset in Tanzania:

@@

We also have seen variability in population trends within species (e.g., two apparently increasing populations and two apparently decreasing populations in Tanzania), so we would want to think a lot more about how we scale up from population to species to capture this variation in a way that is most useful to our question. Here we might realize we want to try to understand the drivers of this variability -- for this we would need to return to Step 1 and reformulate our question, then go through the workflow again, likely adding in potential predictors such as some metric of conservation efforts for that species and or population (e.g., stopping the Ivory trade for elephants, metrics of poaching in different parks) and possibly also country, which may capture differences in efforts related to conservation that are not yet measured. We would do this, returning to Steps 2-5 likely several times until we have a model that is good enough for our purposes here. And then we would likely add in more countries and species.

Alternatively, we may feel this model is good enough for our question and think about how to robustly add in the species and geography we originally excluded.

I might also add a short paragraph about moving away from linear regression and towards models for species abudance (via mixtures I think? I suspect the articles I posted for class have some good ideas) and thinking more about linking to demography models.

And then we would realize we really don't have great data for this question and we would ...

>>>>

At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one sigma for all populations (CHECK HERE: do we really use one sigma for all populations?), how we model population trends (e.g., there are many ways to model abundance) and (relatedly) our linearity assumption. One possibility would be to move to a non-linear regression, and use a more flexible model that would allow population numbers to stabilize, such as shown using a Gompertz model for the wildbeest dataset in Tanzania:

@@

We also have seen variability in population trends within species (e.g., two apparently increasing populations and two apparently decreasing populations in Tanzania), so we would want to think a lot more about how we scale up from population to species to capture this variation in a way that is most useful to our question. For example, may want to think about adding country, which may capture differences in efforts related to conservation and thus predict trends (or we may have better metrics of what drives these trends, but given our global aim, we would need it for most/all populations) and we may want to think about how or if we partially pool species and populations within species (potentially using models other than a normal distribution for pooling). As we address these potential problems, we would likely begin to add in more countries, species, and populations.

We expect scaling up to more countries, species and populations may cause us to think carefully about the data we have in hand versus the question we aimed to answer. While many recent papers treat the LPI data as effectively a well stratified and randomized sample or even effectively a good cencus, it is clearly not. Thus, models that try to deal with such uneven sampling and build in the more of the structure of data, would be far more appropriate. For example, most models for election forecasts now include regularization via partial pooling, post-stratificatiob and other methods designed to adjust for the fact that many polls are 'non-representative.' Such polls may be more similar to LPI data. Our model includes some regularization via partial pooling, but we may need far more of an observational model added to our population demography model to accommodate the uneven sampling. Estimates from such a model may be far more uncertain, but likely also more correct given our combination of question and data.

This may lead us to realize we really don't have great data for this question and we would then need to prioritize better data collection efforts if we truly want good estimates to answer this question.

## The full process

The small workflow we should here may seem like extra work to those who have used it, and it is compared to many ecological workflows in practice. We argue that the extra effort is worth it given the critical questions we are addressing, and also that the process becomes much quicker with practice. The first time may be slow, but the resulting model will be better for it (even if it is still arguably simple, as the one we show here is), but the following times will go faster with likely improved models.
